---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
Coursera Reproducible Research  
Peer Assessment 1  
Jani Savolainen  


## Loading and preprocessing the data
```{r}
fileUrl <- "https://github.com/zavolainen/RepData_PeerAssessment1/raw/master/activity.zip"
fileName <- "activity.zip"

# download data file if it does not exist
if (!file.exists(fileName)) {
        download.file(fileUrl, fileName, mode = "wb")
}

# unzip data file if data directory doesn't exist
dataPath <- "activity"
if (!file.exists(dataPath)) {
        unzip(fileName)
}

data <- read.csv("./activity/activity.csv")

```


## What is mean total number of steps taken per day?
```{r}

# Sum the steps by date and draw a histogram #
aggStepsDay <- aggregate(steps ~ date, data, sum, na.rm=TRUE)
hist(aggStepsDay$steps,  xlab = "Total daily Steps", ylab = "Frequency", main = "", breaks = 25, col = "red", border = "black")

meanStepsDay <- format(as.numeric(mean(aggStepsDay$steps)), scientific = FALSE)
medianStepsDay <- format(as.numeric(median(aggStepsDay$steps)), scientific = FALSE)

```

**MEAN** of the total number of steps taken per day is `r meanStepsDay` and  
**MEDIAN** of the total number of steps taken per day is `r medianStepsDay`.  


## What is the average daily activity pattern?

```{r}

library(ggplot2)

# sum the steos by interval and draw a line graph #
aggStepsInt <- aggregate(steps ~ interval, data = data, FUN = mean, na.rm=TRUE)

stepsIntPlot <- ggplot(data = aggStepsInt, aes(x=interval, y=steps))
stepsIntPlot <- stepsIntPlot + geom_line() + labs(title = "Avgerage Daily Steps by Interval", x = "Interval", y = "Avg. steps per day")
print(stepsIntPlot)

# calculate the interval with most steps #
maxStepsInterval <- aggStepsInt[which(aggStepsInt$steps == max(aggStepsInt$steps)),]
print(maxStepsInterval)

```

On average across all the days in the dataset the 5-minute interval which contains the maximum number of steps is interval `r maxStepsInterval$interval`, where there are about `r maxStepsInterval$steps` steps on average.  

## Imputing missing values

```{r}
# calculate the na's #
missingValuesCount <- length(which(is.na(data$steps)))
missingValuesCount

```

Total number of missing values in the dataset is `r missingValuesCount`.

```{r}

# replace na's in steps column with daily average #
replaceNaWithMean <- transform(data, steps = ifelse(is.na(data$steps), 
                aggStepsInt$steps[match(data$interval, aggStepsInt$interval)], data$steps))
sumReplacedData <- aggregate(steps ~ date, replaceNaWithMean, sum, na.rm=TRUE)

# draw a histogram with the replaced NA's #
hist(sumReplacedData$steps,  xlab = "Total daily Steps (NA's replaced with daily mean)", ylab = "Frequency", main = "", breaks = 25, col = "red", border = "black")

# calculate the new mean and median #
oldmean <- meanStepsDay
newmean <- format(as.numeric(mean(sumReplacedData$steps)), scientific = FALSE)

oldmedian <- medianStepsDay 
newmedian <- format(as.numeric(median(sumReplacedData$steps)), scientific = FALSE)

```

The old mean is `r oldmean` and the new mean with replaced NA's is `r newmean`.  
THe old median is `r oldmedian` and the new median with replaced NA's is `r newmedian`.

Replacing the NA's with daily average number of steps has **no effect on mean** and **very little impact on the median**.

## Are there differences in activity patterns between weekdays and weekends?

```{r}
# collect the name of the weekday for a new column called weekday and assign weekend/weekday for another new column called weekend#
data$date <- as.Date(data$date)
data$weekday <- weekdays(data$date)
data$weekend <- ifelse(data$weekday=="Saturday" | data$weekday=="Sunday", "weekend", "weekday" )

aggStepsIntDate <- aggregate(steps ~ interval + weekend, data = data, FUN = mean, na.rm=TRUE)

# draw a line graph indicating average steps per interval on during the weekdays and weekends #
weekdayWeekendPlot <- ggplot(data = aggStepsIntDate , aes(x = interval , y = steps)) 
weekdayWeekendPlot <- weekdayWeekendPlot +
                        geom_line() +
                        facet_grid(weekend ~.) +
                        xlab("Interval of 5 minutes") +
                        ylab("Average number of steps")

print(weekdayWeekendPlot)

``` 

As we can see from the line plots, during the weekdays and during the weekend the average number of steps in different intervals vary quite a lot.
